---
title: "Demography and Event History Analysis using R"
author: "Göran Broström"
date: '2015-10-01'
output:
  ioslides_presentation:
    fig_height: 2.5
    highlight: tango
  beamer_presentation:
    fig_height: 3
    highlight: tango
    keep_tex: yes
  slidy_presentation:
    highlight: tango
css: style.css
---

## Outline

* The summer course 1-12 June 2015
    + http://capa.ddb.umu.se/rr
    + https://github.com/goranbrostrom/cedar
* Ignoring the <font color="red">matching</font> in the analysis of matched data.
*  What is the Hauck-Donner effect?
* R markdown
* git and gitHub.
    
## Why R?

1. R is *free*.
    * as in free beer
    * as in free research
2. R is open-source
3. R has over 6000 add-on packages
4. R will be the world leading statistical software.
    * More than 7000 packeges on CRAN

## Why RStudio?

Literate programming:

* Dynamic report (paper) writing, programming, and documentation in one.

* Automates *reproducible research*.

* Easy to combine with *version control*.
    * GitHub:   
    [cedar](http://github.com/goranbrostrom/cedar/)
   


## The two-sample problem (example)

```{r whyma, echo=FALSE, fig.height = 6}
set.seed(12135)
X <- rnorm(10, mean = 5, sd = 10)
Y <- X + rnorm(10, mean = 6, sd = 0.2)
plot(1:10, X, col = "red", xlim = c(0, 21), ylim = c(min(X, Y), max(X, Y)),
     xlab = "Index", ylab = "Y")
points(11:20, Y, col = "blue")
```

Statistically significant difference?

## The t-test {.smaller}

\scriptsize

```{r ttest}
t.test(X, Y)
```


## Paired data 


```{r pairda, echo = FALSE, fig.height = 6}
plot(1:10, X, col = "red", xlim = c(0, 21), ylim = c(min(X, Y), max(X, Y)),
     xlab = "", ylab = "Y")
points(11:20, Y, col = "blue")
for (i in 1:10){
    lines(c(i, i + 10), c(X[i], Y[i]), lty = 2)
}
```

## The paired t-test

\scriptsize
```{r pairttest, echo = TRUE}
t.test(X, Y, paired = TRUE)
```

\normalsize

The sign test: p = 2 * .5^10 = 0.00195 (two-sided)

## The 'Hauck-Donner effect' (example)

```{r hdata, fig.height=6,echo=FALSE}
dat = data.frame(T = 1:10, x = c(1:4, 5.02, 4.98, 7:10), event = rep(1, 10))
dat = dat[order(dat$x), ]
with(dat, plot(x, T, type = "b", col = "blue", lty = 2, ylim = c(0, 10)))
```

## Linear regression, code

```{r code, results='hide'}
options(show.signif.stars = FALSE)
fit <- lm(T ~ x, data = dat)
summary(fit)
```

## Linear regression, results

\scriptsize

```{r linres, echo=FALSE}
summary(fit)
```


## ANOVA

```{r anova}
anova(fit)
```

*Note* that the **F test** is appropriate for _linear models_ (with unknown variance)

## Cox regression, 

Wald test:

```{r coxreg, message=FALSE, echo=FALSE}
library(eha)
fit = coxph(Surv(T, event) ~ x, data = dat)
knitr::kable(summary(fit)$coef, digits = 3)
```

LR test:

```{r drop1an, echo = FALSE}
dr = anova(fit)
knitr::kable(dr, digits = 4)
```

## Cox regression, Likelihood Ratio Test

```{r drop1an2}
anova(fit)
```

 *Note* the **Hauck-Donner** effect! 
 
 Strikes in *non-linear* models (not only Cox)!

