---
title: "Demography and Event History Analysis using R"
author: "Göran Broström"
date: '2015-06-18'
output:
  beamer_presentation:
    fig_height: 3
    highlight: tango
    keep_tex: yes
  ioslides_presentation:
    fig_height: 3
    highlight: tango
    logo: figures/umu15.png
  slidy_presentation:
    highlight: tango
css: style.css
---


## Why R?

1. R is *free*.
    * as in free beer
    * as in free research
2. R is open-source
3. R has over 6000 add-on packages
4. R will be the world leading statistical software.

## Why RStudio?

Literate programming (Knuth, 1984):

* Dynamic report (paper) writing, programming, and documentation in one.

* Automates *reproducible research*.

* Easy to combine with *version control*.
    * GitHub:   
    [cedar](http://github.com/goranbrostrom/cedar/)
   


## Statistical issues

or, *Bad habits in Statistics and how to avoid them*.

1. Interactions
    * centering
    * Presentation of results
2. Omitted covariates in PH regression
    * Don't!
3. Causality and matching
4. The Hauck-Donner effect
    * Bann the Wald $p$-values and significance stars!
    
## Interaction

Data from Skellefteå: Old age mortality, 

* ses and 
* birthdate 
* with interaction

```{r interact, echo=FALSE,message=FALSE}
library(eha)
fit0 <- coxph(Surv(enter, exit, event) ~ ses * birthdate, 
             data = mort)
knitr::kable(summary(fit0)$coef[, c(1, 2, 3, 5)], digits = 3)
```

## Center birthdate around 1810

```{r interactcent, echo=TRUE,message=FALSE}
mort$birthdate <- mort$birthdate - 1810
```

```{r iancpr,echo=FALSE}
fit1 <- coxph(Surv(enter, exit, event) ~ ses * birthdate, 
             data = mort)
knitr::kable(summary(fit1)$coef[, c(1, 2, 3, 5)], digits = 3)
knitr::kable(drop1(fit1, test = "Chisq"), digits = 4)
```


## No interaction

```{r nointeractcent, echo=FALSE,message=FALSE}

fit <- coxph(Surv(enter, exit, event) ~ ses + birthdate, 
             data = mort)
knitr::kable(summary(fit)$coef[, c(1, 2, 3, 5)], digits = 3)
knitr::kable(drop1(fit, test = "Chisq"), digits = 4)
```

## Effect of upper vs. lower SES, no centering (log scale)

```{r pres1log4, echo = FALSE,fig.height = 6}
bd <- seq(0, 1820, length = 1000)
low <- fit0$coef[2] * bd
high <- fit0$coef[1] + low + fit0$coef[3] * bd
plot(bd, high - low, type = "l", xlim = c(0, 1820), col = "red", xlab = "Birth Year", ylab = "Log relative risk (upper)", axes = FALSE)
axis(1, at = c(0, 1000, 1800, 1820))
axis(2)
box()
abline(v = 0)
abline(v = c(1800, 1820), lty = 2)
##lines(bd, high, col = "blue")
```

## Effect of upper vs. lower SES, no centering

```{r pres1next, echo = FALSE,fig.height = 6}
bd <- seq(0, 1820, length = 1000)
low <- fit0$coef[2] * bd
high <- fit0$coef[1] + low + fit0$coef[3] * bd
plot(bd, exp(high - low), type = "l", xlim = c(0, 1820), col = "red", xlab = "Birth Year", ylab = "relative risk (upper)", axes = FALSE)
axis(1, at = c(0, 1000, 1800, 1820))
axis(2, at = c(0, 0.5, round(max(exp(high - low)), 2)))
box()
abline(h = 0, v = 0)
abline(v = c(1800, 1820), lty = 2)
##lines(bd, high, col = "blue")
```

## Effect of upper vs. lower SES, with centering (log scale)

```{r pres1log2, echo = FALSE,fig.height = 6}
bd <- seq(1800, 1820, length = 1000)
low <- fit0$coef[2] * bd
high <- fit0$coef[1] + low + fit0$coef[3] * bd
plot(bd, high - low, type = "l", xlim = c(1800, 1820), col = "red", xlab = "Birth Year", ylab = "Log relative risk (upper)", axes = FALSE)
axis(1, at = c(1800, 1810, 1820))
axis(2, at = c(-0.8, round(fit1$coef[1], 2), -0.3))
box()
abline(v = 1810, lty = 2)
abline(h = round(fit1$coef[1], 2), lty = 2)
##lines(bd, high, col = "blue")
```

## Effect of upper vs. lower SES, with centering

```{r pres12, echo = FALSE,fig.height = 6}
bd <- seq(1800, 1820, length = 1000)
low <- fit0$coef[2] * bd
high <- fit0$coef[1] + low + fit0$coef[3] * bd
plot(bd, exp(high - low), type = "l", xlim = c(1800, 1820), col = "red", xlab = "Year", ylab = "relative risk (upper)", axes = FALSE)
axis(1, at = c(1800, 1810, 1820))
axis(2, at = c(0, round(exp(fit1$coef[1]), 2), round(max(exp(high - low)), 2)))
box()
abline(v = 1810, lty = 2)
abline(h = exp(fit1$coef[1]), lty = 2)
##lines(bd, high, col = "blue")
```


<!--
## Effect of upper vs. lower SES, no centering (log scale)

```{r pres1log3, echo = FALSE,fig.height = 6}
bd <- seq(0, 1820, length = 1000)
low <- fit0$coef[2] * bd
high <- fit0$coef[1] + low + fit0$coef[3] * bd
plot(bd, high - low, type = "l", xlim = c(0, 1820), col = "red", xlab = "Year", ylab = "Log relative risk (upper)", axes = FALSE)
axis(1, at = c(0, 1000, 1800, 1820))
axis(2)
box()
abline(v = 0)
abline(v = c(1800, 1820), lty = 2)
##lines(bd, high, col = "blue")
```

## Effect of upper vs. lower SES, no centering

```{r pres1, echo = FALSE,fig.height = 6}
bd <- seq(0, 1820, length = 1000)
low <- fit0$coef[2] * bd
high <- fit0$coef[1] + low + fit0$coef[3] * bd
plot(bd, exp(high - low), type = "l", xlim = c(0, 1820), col = "red", xlab = "Year", ylab = "relative risk (upper)", axes = FALSE)
axis(1, at = c(0, 1000, 1800, 1820))
axis(2, at = c(0, 0.5, max(exp(high - low))))
box()
abline(h = 0, v = 0)
abline(v = c(1800, 1820), lty = 2)
##lines(bd, high, col = "blue")
```

-->

## Omitted covariates, example (simulated data)

```{r simdata, echo=FALSE}
source("R/sim.R")
dat <- sim()
dat$enter <- NULL
names(dat) <- c("time", "death", "gender", "civ")
dat$death <- as.logical(dat$death)
dat$gender <- factor(dat$gender, labels = c("woman", "man"))
dat$civ <- factor(dat$civ, labels = c("married", "unmarried"))
knitr::kable(head(dat))
```

## Summary data

Counts:

```{r sumdat, echo=FALSE}
knitr::kable(with(dat, table(gender, civ)))
```

Mean remaining life times (after 60):

```{r meanexp, echo = FALSE}
T <- matrix(0, ncol = 2, nrow = 2)
T[1, 1] <- mean(dat$time[dat$gender == "woman" & dat$civ =="married"])
T[1, 2] <- mean(dat$time[dat$gender == "woman" & dat$civ =="unmarried"])
T[2, 1] <- mean(dat$time[dat$gender == "man" & dat$civ =="married"])
T[2, 2] <- mean(dat$time[dat$gender == "man" & dat$civ =="unmarried"])
colnames(T) <- c("married", "unmarried")
rownames(T) <- c("woman", "man")
knitr::kable(T, digits = 2)
```

## Analysis (but wrong presentation!)

```{r analy, echo = FALSE, results = 'asis', message=FALSE}
fit.gender <- coxreg(Surv(time, death) ~ gender, data = dat)
fit.both <- coxreg(Surv(time, death) ~ gender + civ, data = dat)
library(stargazer)
stargazer(fit.gender, fit.both, header = FALSE, font.size = "scriptsize")
```

## Simpson's paradox!

(Exponential distribution)

$$
\text{Relative Bias} \approx \frac{(n_{11} \alpha + n_{12})(n_{21} + n_{22})}{(n_{11} + n_{12})(n_{21} \alpha + n_{22})}
$$
```{r relbi, fig.height = 5, echo = FALSE}
source("R/bias.R")
alpha <- seq(0.1, 10, length = 1000)
y <- bias(alpha, exp(1), matrix(c(1000, 2000, 2000, 1000), ncol = 2))
plot(alpha, y, type = "l", col = "blue", ylab = "RB(Hazard ratio)", axes = FALSE)
axis(1, at = c(0.1, 1, 2, 5, 10))
axis(2)
box()
abline(v = 1, lty = 2)
abline(h = 1, lty = 2)
```

## Infant and maternal mortality {.smaller}

Matched, left-truncated data (four first rows). 35 infants whose mothers died. Two controls for each.
```{r sevenp, echo=FALSE,results='asis'}
library(eha)
##library(xtable)
knitr::kable(head(infants[, 1:5], 4))
knitr::kable(head(infants[, 6:11], 4))
```

##  The analysis

Stratify!

```{r strat, echo=TRUE,results='asis'}
fit <- coxreg(Surv(enter, exit, event) ~ mother + 
                  strata(stratum), data = infants)
dr <- drop1(fit, test = "Chisq")
ltx(fit, dr = dr, digits=4)
```

## References

Broström, G. (1987), The influence of mother's mortality on infant mortality. *Scandinavian Journal of Statistics* **14**, 251--261.

Broström, G. (2012). *Event History Analysis with R.* CRC Press/Chapman & Hall, Boca Raton, CA, USA.

## The two-sample problem (example)

```{r whyma, echo=FALSE, fig.height = 6}
set.seed(12135)
X <- rnorm(10, mean = 5, sd = 10)
Y <- X + rnorm(10, mean = 6, sd = 0.2)
plot(1:10, X, col = "red", xlim = c(0, 21), ylim = c(min(X, Y), max(X, Y)),
     xlab = "Index", ylab = "Y")
points(11:20, Y, col = "blue")
```

Statistically significant difference?

## The t-test {.smaller}

\scriptsize

```{r ttest}
t.test(X, Y)
```


## Paired data 


```{r pairda, echo = FALSE, fig.height = 6}
plot(1:10, X, col = "red", xlim = c(0, 21), ylim = c(min(X, Y), max(X, Y)),
     xlab = "", ylab = "Y")
points(11:20, Y, col = "blue")
for (i in 1:10){
    lines(c(i, i + 10), c(X[i], Y[i]), lty = 2)
}
```

## The paired t-test

\scriptsize
```{r pairttest, echo = TRUE}
t.test(X, Y, paired = TRUE)
```

\normalsize

The sign test: p = 2 * .5^10 = 0.00195 (two-sided)

## The 'Hauck-Donner effect' (example)

```{r hdata, fig.height=6, fig.width = 6, echo=FALSE}
dat = data.frame(T = 1:10, x = c(1:4, 5.02, 4.98, 7:10), event = rep(1, 10))
dat = dat[order(dat$x), ]
with(dat, plot(x, T, type = "b", col = "blue", lty = 2, ylim = c(0, 10), xlim = c(0, 10)))
```

## Linear regression, code

```{r code, results='hide'}
options(show.signif.stars = FALSE)
fit <- lm(T ~ x, data = dat)
summary(fit)
```

## Linear regression, results

\scriptsize

```{r linres, echo=FALSE}
summary(fit)
```


## ANOVA

```{r anova}
anova(fit)
```

*Note* that the **F test** is appropriate for _linear models_ (with unknown variance)

## Cox regression, 

Wald test:

```{r coxreg, message=FALSE, echo=FALSE, results = 'asis'}
library(eha)
fit = coxreg(Surv(T, event) ~ x, data = dat)
##knitr::kable(summary(fit)$coef, digits = 3)
ltx(fit)
```

LR test:

\scriptsize

```{r drop1an, echo = FALSE, results = 'asis'}
dr = drop1(fit, test = "Chisq")
##knitr::kable(dr, digits = 4)
ltx(fit, dr = dr, digits = 6)
```


## Cox regression, Likelihood Ratio Test

<!--
```{r drop1an2}
anova(fit)
```
-->

 *Note* the **Hauck-Donner** (1977; *JASA* **72** 851--853) effect! 
 
 Strikes in *non-linear* models (not only Cox)!

## The "graphical"" explanation

```{r grapex, echo = FALSE, fig.height = 7}
source("R/koll2.R")
res = koll2()
```

## Wald and presentation (bad)

````{r bad, echo = FALSE, results = 'asis'}
fit <- coxreg(Surv(enter, exit, event) ~ civ + region, data = oldmort)
ltx(fit)
```

## LR and presentation (good)

````{r good, echo = FALSE, results = 'asis'}
fit <- coxreg(Surv(enter, exit, event) ~ civ + region, data = oldmort)
dr <- drop1(fit, test = "Chisq")
ltx(fit, dr = dr, digits = 4)
```

